---
output: html_document
editor_options: 
  chunk_output_type: inline
---
# Applications

```{r setup method 4, include = FALSE}
knitr::opts_chunk$set(cache = TRUE)
library(tidyverse)
library(readxl)
library(lubridate)
library(ANOVAreplication)
library(multcompView)
library(knitr)
```

The timepoint data was reviously analyzed for headway distributions corresponding to the four separate TSP thresholds for the entire route and T-tests indicated that the means of the headway distributions were not significantly different. However, 
the distributions of the four thresholds were not the same. That brief analysis is shown below.

```{r load 4, echo = FALSE}
#Download data from data file
(df <- read_csv(
  "Data/UVX_Reliability_2019.zip",
  col_types = str_c(rep("c", 24), collapse = "")
  ))
```

```{r my_df 4, echo = FALSE}
(df <- df %>%
  transmute(
    # Route and direction IDs
    route = Route, direction = substr(Direction, 0, 2), 
    trip = Trip,
    timepoint = `Time Point`,
    vehicle = Vehicle,
    
    # Time points
    date = as_date(mdy(NEW_Date)),
    time = as_datetime(str_c(mdy(NEW_Date), " ", DepartureTime)),
    schedule = as_datetime(str_c(mdy(NEW_Date), " ", Schedule, ":00")),
    reliability = time - schedule,
    dwell = as.difftime(str_c("00:", Dwell)),
    travel = as.difftime(Travel),
    
    # change level descriptions for TSP thresholds
    threshold = factor(Threshold, levels = c("OFF", "5", "2", "ON"), 
                       labels = c("No TSP", "5 min", "2 min", "Always"))
  ))
```

```{r headway 4, echo = FALSE}
df <- df %>%
  group_by(direction, timepoint) %>%
  arrange(time, .by_group = TRUE) %>%
  mutate(
    hw_actl = time - lead(time),
    hw_schd = schedule - lead(schedule),
    discrepancy = hw_schd - hw_actl
  )
```

```{r cumdwell 4, echo = FALSE}
df <- df %>%
  group_by(direction, trip, date) %>%
  arrange(time, .by_group = TRUE) %>%
  mutate(
    cumdwell = cumsum(as.numeric(dwell))
  )
```

```{r study-df 4, echo = FALSE}
study_df <- df %>%
  ungroup() %>%
  mutate(
    hour = lubridate::hour(time),
    weekday = lubridate::wday(time)
  ) %>%
  filter(
    !weekday %in% c("Saturday", "Sunday"),
    hour >= 7, hour <= 20
  )

study_df <- study_df[!(study_df$timepoint == "PROVFRST - 1" | study_df$timepoint == "PROVFRST - 2"),]
```

```{r recreate previous analysis, echo = FALSE}
#Show that distributions headway discrepancy grouped by TSP threshold are different
#Create dataframes for each TSP threshold
fivemindf <- filter(study_df, threshold == "5 min")
twomindf <- filter(study_df, threshold == "2 min")
notspdf <- filter(study_df, threshold == "No TSP")
alwaysdf <- filter(study_df, threshold == "Always")

#Create histograms of headway discrepancy for each TSP threshold (update x axes to be same)
ggplot(fivemindf, aes(x = discrepancy / 60)) +
         geom_histogram(binwidth = 2) +
          xlim(-25, 25) +
          labs(x = "Headway Discrepancy (minutes)", title = "Headway Discrepancy Distribution for '5 Min' Threshold")

ggplot(twomindf, aes(x = discrepancy / 60)) +
         geom_histogram(binwidth = 2) +
          xlim(-25, 25) +
          labs(x = "Headway Discrepancy (minutes)", title = "Headway Discrepancy Distribution for '2 Min' Threshold")

ggplot(notspdf, aes(x = discrepancy / 60)) +
         geom_histogram(binwidth = 2) +
          xlim(-25, 25) +
          labs(x = "Headway Discrepancy (minutes)", title = "Headway Discrepancy Distribution for 'No TSP' Threshold")

ggplot(alwaysdf, aes(x = discrepancy / 60)) +
         geom_histogram(binwidth = 2) +
          xlim(-25, 25) +
          labs(x = "Headway Discrepancy (minutes)", title = "Headway Discrepancy Distribution for 'ALWAYS' Threshold")
```

The histograms above show that comparison of headway discrepancy for the respective TSP thresholds considered across the entire bus route yields little visual implication of a difference in distributions. The only noticeable difference between the distributions are the frequencies of headway discrepancies for the respective thresholds due to the varying lengths of time that each was in effect. A one-way ANOVA indicates that the means of these distributions are not significantly different, as shown below.

```{r previous analysis anova, echo = FALSE}

#Convert data to numeric
study_df$reliability <- as.numeric(study_df$reliability)
study_df$dwell <- as.numeric(study_df$dwell)
study_df$travel <- as.numeric(study_df$travel)
study_df$hw_actl <- as.numeric(study_df$hw_actl)
study_df$hw_schd <- as.numeric(study_df$hw_schd)
study_df$discrepancy <- as.numeric(study_df$discrepancy)
study_df$cumdwell <- as.numeric(study_df$cumdwell)

#Use Analysis of Variance to prove that means are not statistically different (t-tests or ANOVA proving Tukey isn't necessary)

lmtsp <- lm(discrepancy ~ threshold, data = study_df)
anova(lmtsp)
```
An *F*-test of the headway discrepancy as a function of TSP thresholds indicates a large *p*-value of 0.2815, which provides no evidence that there is a difference in mean discrepancy between the thresholds when considered in the context of the entire route.

Further investigation of the data may explain differences that are seen between TSP thresholds but that do not significantly change the means of their respective distributions. To investigate the effects of TSP thresholds on headway reliability, a number of things can be done, including splitting up the routes by direction and location and performing additional statistical tests on the data to analyze for variation between groups. A preliminary analysis of the data split in that way is shown below.

## Grouped Distributions

A comparison of headway distributions for separate locations and directions of travel may be insightful into the effects of TSP at different points along the UVX route. For this purpose, the route has been partitioned into the following 12 groups. Groups were chosen to capture groups of riders from four separate sections of the route, as well as to address major intersections where TSP may have more of an influence in one of two directions. In order to provide better comparisons of NB and SB routes, the PROVO groups do not include the stops south of Provo Central Station, since they only appear in the SB trips.

|Group|Starting Station|Ending Station|Direction|Number of Stops|Length|
|----|-----------------|--------------|---------|---------------|------|
|UVU|Orem Central|Main Street|Both|4|2.50 miles|
|UVUNB|Main Street|Orem Central|NB|4|2.50 miles|
|UVUSB|Orem Central|Main Street|SB|4|2.50 miles|
|UNI|Main Street|Byu Stadium|Both|4|2.80 miles|
|UNINB|Byu Stadium|Main Street|NB|4|2.80 miles|
|UNISB|Main Street|Byu Stadium|SB|4|2.80 miles|
|BYU|Byu Stadium|Academy Square|Both|5|2.35 miles|
|BYUNB|Academy Square|Byu Stadium|NB|5|2.35 miles|
|BYUSB|Byu Stadium|Academy Square|SB|5|2.35 miles|
|PROVO|Academy Square|Provo Central|Both|5|1.85 miles|
|PROVONB|Provo Central|Academy Square|NB|5|1.85 miles|
|PROVOSB|Academy Square|Provo Central|SB|5|1.85 miles|


```{r grouped routes, echo = FALSE}

#Make timepoint data a factor
study_df$timepoint <- as.factor(study_df$timepoint)

#Create the UVU group from Orem Central to Main St
UVUvector <- c("OREMFRST", "UVU-CMPS", "LAKEVIEW", "MAIN--ST")
UVU <- study_df %>%
  filter(timepoint %in% UVUvector)

#Create UVU NB and UVU SB groups
UVUNB <- filter(UVU, direction == "NB")
UVUSB <- filter(UVU, direction == "SB")

#Create the UNI group from Main St to BYU Stadium
UNIvector <- c("MAIN--ST", "UNIVPLAC", "2230---N", "BYU-STAD")
UNI <- study_df %>%
  filter(timepoint %in% UNIvector)

#Create the UNI NB and SB groups
UNINB <- filter(UNI, direction == "NB")
UNISB <- filter(UNI, direction == "SB")

#Create the BYU group from BYU Stadium to Academy Square
BYUvector <- c("BYU-STAD", "BYU----N", "BYU-CMPS", "JOA-QUIN", "ACADSQUR")
BYU <- study_df %>%
  filter(timepoint %in% BYUvector)

#Create the BYU NB and SB groups
BYUNB <- filter(BYU, direction == "NB")
BYUSB <- filter(BYU, direction == "SB")

#Create the PROVO group from Academy Square to East Bay South
PROVOvector <- c("ACADSQUR", "300----N", "CNTR--ST", "400----S", "PROVFRST")
PROVO <- study_df %>%
  filter(timepoint %in% PROVOvector)

#Create the PROVO NB and SB groups
PROVONB <- filter(PROVO, direction == "NB")
PROVOSB <- filter(PROVO, direction == "SB")
```

Histograms for the separate groups will be created for initial visualization of the data, so as to inspect for differences among the groups. The first four histograms show the UVU, UNI, BYU, and PROVO groups, respectively. The second four histograms show the same groups for NB trips, and the last four show the same groups for the SB trips.

```{r group histograms, echo = FALSE}

#Create UVU histogram
UVU %>%
  ungroup() %>%
  mutate(discrepancy = as.numeric(abs(discrepancy) + 1)) %>% # +1 for log(0)
  ggplot(aes(x = discrepancy / 60)) + # convert seconds to minutes
  geom_histogram() +
  scale_x_log10() +
  xlab("Absolute discrepancy between scheduled and actual headways [minutes]") +
  ggtitle("UVU Route")

#Create UNI histogram
UNI %>%
  ungroup() %>%
  mutate(discrepancy = as.numeric(abs(discrepancy) + 1)) %>% # +1 for log(0)
  ggplot(aes(x = discrepancy / 60)) + # convert seconds to minutes
  geom_histogram() +
  scale_x_log10() +
  xlab("Absolute discrepancy between scheduled and actual headways [minutes]") +
  ggtitle("UNI Route")

#Create BYU histogram
BYU %>%
  ungroup() %>%
  mutate(discrepancy = as.numeric(abs(discrepancy) + 1)) %>% # +1 for log(0)
  ggplot(aes(x = discrepancy / 60)) + # convert seconds to minutes
  geom_histogram() +
  scale_x_log10() +
  xlab("Absolute discrepancy between scheduled and actual headways [minutes]") +
  ggtitle("BYU Route")

#Create PROVO histogram
PROVO %>%
  ungroup() %>%
  mutate(discrepancy = as.numeric(abs(discrepancy) + 1)) %>% # +1 for log(0)
  ggplot(aes(x = discrepancy / 60)) + # convert seconds to minutes
  geom_histogram() +
  scale_x_log10() +
  xlab("Absolute discrepancy between scheduled and actual headways [minutes]") +
  ggtitle("PROVO Route")

#Create UVUNB histogram
UVUNB %>%
  ungroup() %>%
  mutate(discrepancy = as.numeric(abs(discrepancy) + 1)) %>% # +1 for log(0)
  ggplot(aes(x = discrepancy / 60)) + # convert seconds to minutes
  geom_histogram() +
  scale_x_log10() +
  xlab("Absolute discrepancy between scheduled and actual headways [minutes]") +
  ggtitle("UVU NB Route")

#Create UNINB histogram
UNINB %>%
  ungroup() %>%
  mutate(discrepancy = as.numeric(abs(discrepancy) + 1)) %>% # +1 for log(0)
  ggplot(aes(x = discrepancy / 60)) + # convert seconds to minutes
  geom_histogram() +
  scale_x_log10() +
  xlab("Absolute discrepancy between scheduled and actual headways [minutes]") +
  ggtitle("UNI NB Route")

#Create BYUNB histogram
BYUNB %>%
  ungroup() %>%
  mutate(discrepancy = as.numeric(abs(discrepancy) + 1)) %>% # +1 for log(0)
  ggplot(aes(x = discrepancy / 60)) + # convert seconds to minutes
  geom_histogram() +
  scale_x_log10() +
  xlab("Absolute discrepancy between scheduled and actual headways [minutes]") +
  ggtitle("BYU NB Route")

#Create PROVONB histogram
PROVONB %>%
  ungroup() %>%
  mutate(discrepancy = as.numeric(abs(discrepancy) + 1)) %>% # +1 for log(0)
  ggplot(aes(x = discrepancy / 60)) + # convert seconds to minutes
  geom_histogram() +
  scale_x_log10() +
  xlab("Absolute discrepancy between scheduled and actual headways [minutes]") +
  ggtitle("PROVO NB Route")

#Create UVUSB histogram
UVUSB %>%
  ungroup() %>%
  mutate(discrepancy = as.numeric(abs(discrepancy) + 1)) %>% # +1 for log(0)
  ggplot(aes(x = discrepancy / 60)) + # convert seconds to minutes
  geom_histogram() +
  scale_x_log10() +
  xlab("Absolute discrepancy between scheduled and actual headways [minutes]") +
  ggtitle("UVU SB Route")

#Create UNISB histogram
UNISB %>%
  ungroup() %>%
  mutate(discrepancy = as.numeric(abs(discrepancy) + 1)) %>% # +1 for log(0)
  ggplot(aes(x = discrepancy / 60)) + # convert seconds to minutes
  geom_histogram() +
  scale_x_log10() +
  xlab("Absolute discrepancy between scheduled and actual headways [minutes]") +
  ggtitle("UNI SB Route")

#Create BYUSB histogram
BYUSB %>%
  ungroup() %>%
  mutate(discrepancy = as.numeric(abs(discrepancy) + 1)) %>% # +1 for log(0)
  ggplot(aes(x = discrepancy / 60)) + # convert seconds to minutes
  geom_histogram() +
  scale_x_log10() +
  xlab("Absolute discrepancy between scheduled and actual headways [minutes]") +
  ggtitle("BYU SB Route")

#Create PROVOSB histogram
PROVOSB %>%
  ungroup() %>%
  mutate(discrepancy = as.numeric(abs(discrepancy) + 1)) %>% # +1 for log(0)
  ggplot(aes(x = discrepancy / 60)) + # convert seconds to minutes
  geom_histogram() +
  scale_x_log10() +
  xlab("Absolute discrepancy between scheduled and actual headways [minutes]") +
  ggtitle("PROVO SB Route")
```

Inspection of the histograms indicates that the distributions of headway discrepancies for the groups tends to be around 1 minute. There is some variation between the groups, though the distributions appear to be somewhat normal. A table of group means and standard deviations is given below.

```{r groups df, echo = FALSE}
#Collect the discrepancy column from each group dataframe as a vector
UVUdisc <- as.vector(UVU$discrepancy)
UVUNBdisc <- as.vector(UVUNB$discrepancy)
UVUSBdisc <- as.vector(UVUSB$discrepancy)
UNIdisc <- as.vector(UNI$discrepancy)
UNINBdisc <- as.vector(UNINB$discrepancy)
UNISBdisc <- as.vector(UNISB$discrepancy)
BYUdisc <- as.vector(BYU$discrepancy)
BYUNBdisc <- as.vector(BYUNB$discrepancy)
BYUSBdisc <- as.vector(BYUSB$discrepancy)
PROVOdisc <- as.vector(PROVO$discrepancy)
PROVONBdisc <- as.vector(PROVONB$discrepancy)
PROVOSBdisc <- as.vector(PROVOSB$discrepancy)

#Create vectors for each of the factor levels (each of the 12 routes)

UVUroute <- rep("UVU", 190149)
UVUNBroute <- rep("UVUNB", 96854)
UVUSBroute <- rep("UVUSB", 93295)
UNIroute <- rep("UNI", 202718)
UNINBroute <- rep("UNINB", 100875)
UNISBroute <- rep("UNISB", 101843)
BYUroute <- rep("BYU", 253374)
BYUNBroute <- rep("BYUNB", 124976)
BYUSBroute <- rep("BYUSB", 128398)
PROVOroute <- rep("PROVO", 217793)
PROVONBroute <- rep("PROVONB", 115358)
PROVOSBroute <- rep("PROVOSB", 102435)
              
#combine vectors into dataframes for comparing discrepancies (1- compare full routes, 2- compare NB routes, 3- compare SB routes)

#Full routes dataframe
Fullroutes <- as.data.frame(cbind(c(UVUroute, UNIroute, BYUroute, PROVOroute), c(UVUdisc, UNIdisc, BYUdisc, PROVOdisc))) %>% 
  rename(Route = V1, Discrepancy = V2)
Fullroutes$Discrepancy <- as.numeric(Fullroutes$Discrepancy)
Fullroutes$Route <- as.factor(Fullroutes$Route)

#NB routes dataframe
NBroutes <- as.data.frame(cbind(c(UVUNBroute, UNINBroute, BYUNBroute, PROVONBroute), c(UVUNBdisc, UNINBdisc, BYUNBdisc, PROVONBdisc))) %>% 
  rename(Route = V1, Discrepancy = V2)
NBroutes$Discrepancy <- as.numeric(NBroutes$Discrepancy)
NBroutes$Route <- as.factor(NBroutes$Route)

#SB routes dataframe
SBroutes <- as.data.frame(cbind(c(UVUSBroute, UNISBroute, BYUSBroute, PROVOSBroute), c(UVUSBdisc, UNISBdisc, BYUSBdisc, PROVOSBdisc))) %>% 
  rename(Route = V1, Discrepancy = V2)
SBroutes$Discrepancy <- as.numeric(SBroutes$Discrepancy)
SBroutes$Route <- as.factor(SBroutes$Route)
```

```{r groups stats table, echo = FALSE}
#get the means and standard deviations for each group to display, and reorder according to route
Fullmeans <- aggregate(data = Fullroutes, Discrepancy ~ Route, FUN = mean)
a <- c("PROVO", "BYU", "UNI", "UVU")
Fullmeans <- Fullmeans %>%
  slice(match(a, Route))
NBmeans <-  aggregate(data = NBroutes, Discrepancy ~ Route, FUN = mean)
b <- c("PROVONB", "BYUNB", "UNINB", "UVUNB")
NBmeans <- NBmeans %>%
  slice(match(b, Route))
SBmeans <- aggregate(data = SBroutes, Discrepancy ~ Route, FUN = mean)
c <- c("PROVOSB", "BYUSB", "UNISB", "UVUSB")
SBmeans <- SBmeans %>%
  slice(match(c, Route))
Fullsd <- aggregate(data = Fullroutes, Discrepancy ~ Route, FUN = sd)
Fullsd <- Fullsd %>%
  slice(match(a, Route))
NBsd <- aggregate(data = NBroutes, Discrepancy ~ Route, FUN = sd)
NBsd <- NBsd %>%
  slice(match(b, Route))
SBsd <- aggregate(data = SBroutes, Discrepancy ~ Route, FUN = sd)
SBsd <- SBsd %>%
  slice(match(c, Route))

#Create a dataframe for the means and sd's of each group
Groupmeans <- as.data.frame(rbind(Fullmeans, NBmeans, SBmeans))
groupsds <- as.vector(c(Fullsd$Discrepancy, NBsd$Discrepancy, SBsd$Discrepancy))
Groupsummary <- cbind(Groupmeans, groupsds)
Groupsummary <- rename(Groupsummary, "Mean" = Discrepancy, "SD" = groupsds)

#create a nice table for the group summary
kable(Groupsummary)
```

The results of splitting up the trips by location are expected. There is not a large variation of discrepancies when routes are compared with both their northbound and southbound trips included. This is likely due to the trend seen in the northbound and southbound groups. The mean discrepancy increases in the northbound trips as the buses travel from the start of the trip in Provo towards the end of the trip near UVU. The same trend is seen in the southbound trips from UVU to Provo. The standard deviations of headway discrepancies behave the same way. This indicates the cumulative nature of headway discrepancy. As buses progress along their routes, irregularities tend to compound. This is may be due to problems such as the bunching phenomenon discussed in the literature.

By means of data visualization and exploration, we have determined that the data is organized in a way that meets assumptions for further statistical tests. We are specifically interested in the distributions of headway discrepancy for the four respective TSP thresholds. Initial analysis above indicated that there is no statistical difference between mean headway discrepancy for the respective thresholds when considered in the context of the entire bus route. However, we may be able to evaluate certain locations along the route where TSP thresholds do affect the headway discrepancy. If statistical differences do appear, this may indicate that TSP thresholds do play a role for certain portions of the UVX route.


## statistical Tests on Grouped Distributions

For each of the 12 partitioned groups of the UVX route, a boxplot comparing headway discrepancy by TSP threshold will be shown along with an ANOVA table. The differences between these groups are difficult to determine based off of visuals alone, but the *F*-tests can help determine which, if any, differences are significant. If significant differences are observed, then the Tukey-Kramer post-hoc procedure for multiple comparisons of unplanned differences will be used to evaluate them. From those analyses, conclusions will be made about the effect of TSP thresholds on sections of the UVX route.The results of F-tests run to compare TSP thresholds within the partitioned groups are shown below.

```{r boxplot on PROVO, echo = FALSE}
#Remove rows where threshold = "NA" from the PROVO group
PROVO <- PROVO[!is.na(PROVO$threshold),]

#Create a boxplot of headway discrepancies by TSP for the PROVO group
PROVO %>%
  ungroup() %>%
  mutate(discrepancy = as.numeric(abs(discrepancy) + 1)) %>% # +1 for log(0)
  ggplot(aes(x = log(discrepancy / 60), color = threshold)) + # convert seconds to minutes
  geom_boxplot() +
  xlab("Log Minutes") +
  theme(axis.ticks.y = element_blank(), axis.text.y = element_blank()) +
  ggtitle("Comparitive Boxplot of TSP Thresholds for Provo Trips (NB and SB)")
```

```{r F-test on PROVO, echo = FALSE}
#Run F-test on Provo group
PROVOlm <- lm(discrepancy ~ threshold, data = PROVO)
anova(PROVOlm)
```

```{r boxplot on BYU, echo = FALSE}
#Remove rows where threshold = "NA" from the BYU group
BYU <- BYU[!is.na(BYU$threshold),]

#Create a boxplot of headway discrepancies by TSP for the BYU group
BYU %>%
  ungroup() %>%
  mutate(discrepancy = as.numeric(abs(discrepancy) + 1)) %>% # +1 for log(0)
  ggplot(aes(x = log(discrepancy / 60), color = threshold)) + # convert seconds to minutes
  geom_boxplot() +
  xlab("Log Minutes") +
  theme(axis.ticks.y = element_blank(), axis.text.y = element_blank()) +
  ggtitle("Comparitive Boxplot of TSP Thresholds for BYUTrips (NB and SB)")
```

```{r F-test on BYU, echo = FALSE}
#Run F-test on BYU group
BYUlm <- lm(discrepancy ~ threshold, data = BYU)
anova(BYUlm)
```

```{r boxplot on UNI, echo = FALSE}
#Remove rows where threshold = "NA" from the UNI group
UNI <- UNI[!is.na(UNI$threshold),]

#Create a boxplot of headway discrepancies by TSP for the UNI group
UNI %>%
  ungroup() %>%
  mutate(discrepancy = as.numeric(abs(discrepancy) + 1)) %>% # +1 for log(0)
  ggplot(aes(x = log(discrepancy / 60), color = threshold)) + # convert seconds to minutes
  geom_boxplot() +
  xlab("Log Minutes") +
  theme(axis.ticks.y = element_blank(), axis.text.y = element_blank()) +
  ggtitle("Comparitive Boxplot of TSP Thresholds for UNI Trips (NB and SB)")
```

```{r F-test on UNI, echo = FALSE}
UNIlm <- lm(discrepancy ~ threshold, data = UNI)
anova(UNIlm)
```

```{r boxplot on UVU, echo = FALSE}
#Remove rows where threshold = "NA" from the UVU group
UVU <- UVU[!is.na(UVU$threshold),]

#Create a boxplot of headway discrepancies by TSP for the UVU group
UVU %>%
  ungroup() %>%
  mutate(discrepancy = as.numeric(abs(discrepancy) + 1)) %>% # +1 for log(0)
  ggplot(aes(x = log(discrepancy / 60), color = threshold)) + # convert seconds to minutes
  geom_boxplot() +
  xlab("Log Minutes") +
  theme(axis.ticks.y = element_blank(), axis.text.y = element_blank()) +
  ggtitle("Comparitive Boxplot of TSP Thresholds for UVU Trips (NB and SB)")
```

```{r F-test on UVU, echo = FALSE}
UVUlm <- lm(discrepancy ~ threshold, data = UVU)
anova(UVUlm)
```

visualization of the cumulative headway discrepancies for the respective TSP thresholds shows little variation between the locations, and nothing that strongly indicates differences between the effects of TSP thresholds. Results from the *F*-tests indicate no evidence of a difference between mean headway discrepancies based on TSP threshold in any of the above groups, as indicated by the high reported *p*-values (0.5722 for the PROVO group, 0.8082 for the BYU group, 0.9808 for the UNI group, and 0.9524 for the UVU group). This finding is not surprising considering the cumulative nature of the headway discrepancy when a group is near the start or end of a trip. The NB and SB discrepancies likely mitigate each other to a certain extent. We may find different results when analysis is run on the NB or SB groups.


The results of F-tests run to compare TSP thresholds within the partitioned NB groups are shown below.

```{r boxplot on PROVONB, echo = FALSE}
#Remove rows where threshold = "NA" from the PROVONB group
PROVONB <- PROVONB[!is.na(PROVONB$threshold),]

#Create a boxplot of headway discrepancies by TSP for the PROVONB group
PROVONB %>%
  ungroup() %>%
  mutate(discrepancy = as.numeric(abs(discrepancy) + 1)) %>% # +1 for log(0)
  ggplot(aes(x = log(discrepancy / 60), color = threshold)) + # convert seconds to minutes
  geom_boxplot() +
  xlab("Log Minutes") +
  theme(axis.ticks.y = element_blank(), axis.text.y = element_blank()) +
  ggtitle("Comparitive Boxplot of TSP Thresholds for PROVO NB Trips")
```

```{r F-test on PROVONB, echo = FALSE}
PROVONBlm <- lm(discrepancy ~ threshold, data = PROVONB)
anova(PROVONBlm)
```

```{r boxplot on BYUNB, echo = FALSE}
#Remove rows where threshold = "NA" from the BYUNB group
BYUNB <- BYUNB[!is.na(BYUNB$threshold),]

#Create a boxplot of headway discrepancies by TSP for the BYUNB group
BYUNB %>%
  ungroup() %>%
  mutate(discrepancy = as.numeric(abs(discrepancy) + 1)) %>% # +1 for log(0)
  ggplot(aes(x = log(discrepancy / 60), color = threshold)) + # convert seconds to minutes
  geom_boxplot() +
  xlab("Log Minutes") +
  theme(axis.ticks.y = element_blank(), axis.text.y = element_blank()) +
  ggtitle("Comparitive Boxplot of TSP Thresholds for BYU NB Trips")
```

```{r F-test on BYUNB, echo = FALSE}
BYUNBlm <- lm(discrepancy ~ threshold, data = BYUNB)
anova(BYUNBlm)
```

```{r boxplot on UNINB, echo = FALSE}
#Remove rows where threshold = "NA" from the UNINB group
UNINB <- UNINB[!is.na(UNINB$threshold),]

#Create a boxplot of headway discrepancies by TSP for the UNINB group
UNINB %>%
  ungroup() %>%
  mutate(discrepancy = as.numeric(abs(discrepancy) + 1)) %>% # +1 for log(0)
  ggplot(aes(x = log(discrepancy / 60), color = threshold)) + # convert seconds to minutes
  geom_boxplot() +
  xlab("Log Minutes") +
  theme(axis.ticks.y = element_blank(), axis.text.y = element_blank()) +
  ggtitle("Comparitive Boxplot of TSP Thresholds for UNI NB Trips")
```

```{r F-test on UNINB, echo = FALSE}
UNINBlm <- lm(discrepancy ~ threshold, data = UNINB)
anova(UNINBlm)
```

```{r Cumulative Headway Discrepancy on UVUNB, echo = FALSE}
#Remove rows where threshold = "NA" from the UVUNB group
UVUNB <- UVUNB[!is.na(UVUNB$threshold),]

#Create a boxplot of headway discrepancies by TSP for the UVUNB group
UVUNB %>%
  ungroup() %>%
  mutate(discrepancy = as.numeric(abs(discrepancy) + 1)) %>% # +1 for log(0)
  ggplot(aes(x = log(discrepancy / 60), color = threshold)) + # convert seconds to minutes
  geom_boxplot() +
  xlab("Log Minutes") +
  theme(axis.ticks.y = element_blank(), axis.text.y = element_blank()) +
  ggtitle("Comparitive Boxplot of TSP Thresholds for UVU NB Trips")
```

```{r F-test on UVUNB, echo = FALSE}
UVUNBlm <- lm(discrepancy ~ threshold, data = UVUNB)
anova(UVUNBlm)
```

The results of F-tests run to compare TSP thresholds within the partitioned SB groups are shown below.

```{r boxplot on PROVOSB, echo = FALSE}
#Remove rows where threshold = "NA" from the PROVOSB group
PROVOSB <- PROVOSB[!is.na(PROVOSB$threshold),]

#Create a boxplot of headway discrepancies by TSP for the PROVOSB group
PROVOSB %>%
  ungroup() %>%
  mutate(discrepancy = as.numeric(abs(discrepancy) + 1)) %>% # +1 for log(0)
  ggplot(aes(x = log(discrepancy / 60), color = threshold)) + # convert seconds to minutes
  geom_boxplot() +
  xlab("Log Minutes") +
  theme(axis.ticks.y = element_blank(), axis.text.y = element_blank()) +
  ggtitle("Comparitive Boxplot of TSP Thresholds for PROVO SB Trips")
```

```{r F-test on PROVOSB, echo = FALSE}
PROVOSBlm <- lm(discrepancy ~ threshold, data = PROVOSB)
anova(PROVOSBlm)
```

```{r boxplot on BYUSB, echo = FALSE}
#Remove rows where threshold = "NA" from the BYUSB group
BYUSB <- BYUSB[!is.na(BYUSB$threshold),]

#Create a boxplot of headway discrepancies by TSP for the BYUSB group
BYUSB %>%
  ungroup() %>%
  mutate(discrepancy = as.numeric(abs(discrepancy) + 1)) %>% # +1 for log(0)
  ggplot(aes(x = log(discrepancy / 60), color = threshold)) + # convert seconds to minutes
  geom_boxplot() +
  xlab("Log Minutes") +
  theme(axis.ticks.y = element_blank(), axis.text.y = element_blank()) +
  ggtitle("Comparitive Boxplot of TSP Thresholds for BYU SB Trips")
```

```{r F-test on BYUSB, echo = FALSE}
BYUSBlm <- lm(discrepancy ~ threshold, data = BYUSB)
anova(BYUSBlm)
```

```{r boxplot on UNISB, echo = FALSE}
#Remove rows where threshold = "NA" from the UNISB group
UNISB <- UNISB[!is.na(UNISB$threshold),]

#Create a boxplot of headway discrepancies by TSP for the UNISB group
UNISB %>%
  ungroup() %>%
  mutate(discrepancy = as.numeric(abs(discrepancy) + 1)) %>% # +1 for log(0)
  ggplot(aes(x = log(discrepancy / 60), color = threshold)) + # convert seconds to minutes
  geom_boxplot() +
  xlab("Log Minutes") +
  theme(axis.ticks.y = element_blank(), axis.text.y = element_blank()) +
  ggtitle("Comparitive Boxplot of TSP Thresholds for UNI SB Trips")
```

```{r F-test on UNISB, echo = FALSE}
UNISBlm <- lm(discrepancy ~ threshold, data = UNISB)
anova(UNISBlm)
```

```{r boxplot on UVUSB, echo = FALSE}
#Remove rows where threshold = "NA" from the UVUSB group
UVUSB <- UVUSB[!is.na(UVUSB$threshold),]

#Create a boxplot of headway discrepancies by TSP for the UVUSB group
UVUSB %>%
  ungroup() %>%
  mutate(discrepancy = as.numeric(abs(discrepancy) + 1)) %>% # +1 for log(0)
  ggplot(aes(x = log(discrepancy / 60), color = threshold)) + # convert seconds to minutes
  geom_boxplot() +
  xlab("Log Minutes") +
  theme(axis.ticks.y = element_blank(), axis.text.y = element_blank()) +
  ggtitle("Comparitive Boxplot of TSP Thresholds for UVU SB Trips")
```

```{r F-test on UVUSB, echo = FALSE}
UVUSBlm <- lm(discrepancy ~ threshold, data = UVUSB)
anova(UVUSBlm)
```

The same pattern of cumulative headway discrepancy is seen by comparison of the routes, with the mean headway discrepancy slightly increasing towards the end of each respective NB or SB route. However, results of the *F*-tests on the groups partitioned by both location and direction were similar to those of the groups partitioned by location alone. A table of *p*-values for all twelve groups is shown below.

|Route Group|*p*-value|
|-----------|---------|
|PROVO|0.5722|
|BYU|0.8082|
|UNI|0.9808|
|UVU|0.9524|
|PROVO (NB)|0.7539|
|BYU (NB)|0.7966|
|UNI (NB)|0.9696|
|UVU (NB)|0.8981|
|PROVO (SB)|0.6013|
|BYUSB (SB)|0.8614|
|UNISB (SB)|0.9675|
|UVUSB (SB)|0.9644|


In none of the groups is there any evidence of a difference between mean headway discrepancy when accounting for the four different TSP thresholds. Even in the PROVO group, which had the lowest *p*-value, the differences between TSP thresholds is insignificant. Estimates of those differences are shown in the following Tukey-Kramer post-hoc test. There is no indication of differences in headway discrepancy for any comparison of TSP thresholds, as all *p*-values reported are all very high. This is consistent across each of the 12 groups, and this analysis is shown here only for reference. 

```{r example Tukey on PROVO group, echo = FALSE}
#Perform a Tukey-Kramer test of multiple comparisons on the PROVO group
AOVPROVO <- aov(PROVOlm)
TukeyPROVO <- TukeyHSD(AOVPROVO)
TukeyPROVO
```
This analysis has shown no evidence that TSP thresholds provide a significant effect on the headway distribution of buses belonging to the UVX BRT system when accounting for location and direction alone. There may be a number of confounding factors not included in this analysis. For example, variability in ridership may affect dwell times and the ability for buses to meet acceptable headways without triggering TSP in certain locations. A simple comparison of cumulative headway distribution by trip direction may indicate this possible effect. The cumulative headway discrepancy plots below show differences between mean headway discrepancies of the UVU, UNI, BYU, or PROVO groups when compared to their respective two-way, NB, or SB counterparts.

```{r cumulative headway discrepancy, echo = FALSE}
#Create a model and perform the F test for the Fullroutes dataframe
Fullrouteslm <- lm(Discrepancy ~ Route, data = Fullroutes)

#Create a model and perform the F test for the NBroutes dataframe
NBrouteslm <- lm(Discrepancy ~ Route, data = NBroutes)

#Create a model and perform the F test for the SBroutes dataframe
SBrouteslm <- lm(Discrepancy ~ Route, data = SBroutes)

#Compare the Full routes
Fullroutes %>%
  ungroup() %>%
  mutate(Discrepancy = as.numeric(abs(Discrepancy) + 1)) %>% # +1 for log(0)
  ggplot(aes(x = Discrepancy / 60, color = Route)) + # convert seconds to minutes
  stat_ecdf() +
  scale_x_log10() +
  xlab("Cumulative discrepancy between scheduled and actual headways [minutes]") +
  ggtitle("Full Routes")

#Compare the NB routes
NBroutes %>%
  ungroup() %>%
  mutate(Discrepancy = as.numeric(abs(Discrepancy) + 1)) %>% # +1 for log(0)
  ggplot(aes(x = Discrepancy / 60, color = Route)) + # convert seconds to minutes
  stat_ecdf() +
  scale_x_log10() +
  xlab("Cumulative discrepancy between scheduled and actual headways [minutes]") +
  ggtitle("NB Routes")

#Compare the SB routes
SBroutes %>%
  ungroup() %>%
  mutate(Discrepancy = as.numeric(abs(Discrepancy) + 1)) %>% # +1 for log(0)
  ggplot(aes(x = Discrepancy / 60, color = Route)) + # convert seconds to minutes
  stat_ecdf() +
  scale_x_log10() +
  xlab("Cumulative discrepancy between scheduled and actual headways [minutes]") +
  ggtitle("SB Routes")
```

It can easily be seen from the plots that in combining the NB and SB trips for each group (PROVO, BYU, UNI, or UVU), the discrepancies do not vary much, likely due to the opposite effect of compounding irregularities from the trip origin. However, there are some differences between the NB and SB groups that are worth noting. In the plot of the NB routes, there appears to be a fairly even distribution of headway discrepancies along the route. No one group appears to be more influential to the discrepancy, but the discrepancy does increase between all sections when buses travel from Provo to UVU. However, in the plot of the SB routes, the cumulative headway distribution varies less between the UNI, BYU, and PROVO sections. This smaller variance demonstrates that headway deviations are not particularly exacerbated during that portion of the bus trips, which could be due to lower demand in the SB trips past the BYU area, as much of the ridership is likely students traveling between BYU and UVU. Further analysis of the relationships between TSP, location, and dwell time could provide more meaningful insights.

Other factors that could be influential to the data include time of year and time of day. Where this analysis was performed considering the geospatial aspects of the timepoint data, it did not consider geotemporal aspects of the BRT system. Time of year was somewhat included in the analysis due to the fact that the different TSP thresholds were in effect at different times of the year. It is possible that TSP thresholds could be more significant during peak or off-peak periods of the day, and this possibility was ignored in the data analysis presented here.

